---
title: "CLL_scRNASeq_seurat"
author: "ASG"
date: "7/9/2021"
output: github_document
toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chronic lymphocytic leukemia single cell RNA-Seq data in two conditions: control & treatment.

The sample is primary CLL tumor from patient grown under two conditions: control: includes fibroblasts and treatment: includes fibroblasts and growth factor.

Some info on the 10X data from the two conditions can be found:
[link] ~/scRNASeq_SR1/data/control/con1.html and here:[link] ~/scRNASeq_SR1/data/treat1/treat1.html. This includes summary and analysis using Cellranger 6.0.1

## loading needed packages
```{r include=FALSE}
library(Seurat)
library(tidyverse)
library(Matrix)
library(patchwork)
```

## reading in data
The data provided is in the "data" folder in this project. Consists of 2 sub-folders representing the two conditions. 

```{r}
# load the cll_control data
cll_cntrl_data <- Read10X("./data/control/con1-filtered_feature_bc_matrix/")
# initializing the Seurat object with raw control data
cll_control <- CreateSeuratObject(counts = cll_cntrl_data, project = "cll", min.cells = 3, min.features = 200)
cll_control
```

```{r}
# load the cll_treatment data
cll_treat1_data <- Read10X("./data/treat1/treat1-filtered_feature_bc_matrix/")
# initializing the Seurat object with raw treatment data (non-normalized)
cll_treatment <- CreateSeuratObject(counts = cll_treat1_data, project = "cll", min.cells = 3, min.features = 200)
cll_treatment
```
## QC metrics and filtering control data based on QC
```{r}
# QC for first 5 cells in control
head(cll_control@meta.data, 5)
```


